---
title: "CITEseq-HIVE clusterProfiler Analysis"
format: html
editor: visual
---

## Description

### Clear console

```{r, output=FALSE}
ls()
rm(list=ls())
```

### Set output directory

```{r}
dir_save <- "output/dataset_full/"
```

### Load libraries

```{r, message = FALSE}
# CHECK THIS CONTAINS ONLY NECESSARY LIBRARIES
library(SingleCellExperiment)
library(Seurat)
library(SeuratObject)
library(dplyr)
library(clustree)
library(ggpubr)
library(patchwork)
library(magrittr)
library(HGNChelper)
library(openxlsx)
library(multtest)
library(metap)
library(ggplot2)
library(cowplot)
library(ComplexHeatmap)
library(magick)
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(DOSE)
library(ggplot2)
library(enrichplot)
```

### Load the .Rdata file generated after RNA clustering only by CITEseqHIVE_Seurat_ClusterIDcells.qmd

```{r}
load("output/dataset_full/Seurat_ClusterIDCellsRNAonly_obj.Rdata")
obj

### Confirm Seruat object contains assays for RNA and ADT
Assays(obj)

### Set default assay to RNA
DefaultAssay(obj) <- "RNA"
```

### Generate subset objs to use for comparisons

```{r}
obj_D0 <- subset(obj, subset = VISIT %in% c("D0"))
obj_D37 <- subset(obj, subset = VISIT %in% c("D37"))
obj_M72 <- subset(obj, subset = STIM %in% c("M72"))
obj_UNSTIM <- subset(obj, subset = STIM %in% c("UNSTIM"))

# Progenitor cells and HSC/MPP cells left out

obj_D0_Lymphoid <- subset(obj_D0, subset = sctype_classification %in% c("Pro-B cells", "Pre-B cells", "Naive B cells", "Memory B cells", "Plasma B cells", "Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_D37_Lymphoid <- subset(obj_D37, subset = sctype_classification %in% c("Pro-B cells", "Pre-B cells", "Naive B cells", "Memory B cells", "Plasma B cells", "Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_M72_Lymphoid <- subset(obj_M72, subset = sctype_classification %in% c("Pro-B cells", "Pre-B cells", "Naive B cells", "Memory B cells", "Plasma B cells", "Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_UNSTIM_Lymphoid <- subset(obj_UNSTIM, subset = sctype_classification %in% c("Pro-B cells", "Pre-B cells", "Naive B cells", "Memory B cells", "Plasma B cells", "Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_Lymphoid <- subset(obj, subset = sctype_classification %in% c("Pro-B cells", "Pre-B cells", "Naive B cells", "Memory B cells", "Plasma B cells", "Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols

obj_D0_Myeloid <- subset(obj_D0, subset = sctype_classification %in% c("Eosinophils", "Neutrophils", "Basophils", "Mast cells", "Classical Monocytes", "Non-classical monocytes", "Intermediate monocytes", "Macrophages", "Megakaryocyte", "Erythroid-like and erythroid precursor cells", "Myeloid Dendritic cells", "Plasmacytoid Dendritic cells", "Granulocytes", "ISG expressing immune cells"))
obj_D37_Myeloid <- subset(obj_D37, subset = sctype_classification %in% c("Eosinophils", "Neutrophils", "Basophils", "Mast cells", "Classical Monocytes", "Non-classical monocytes", "Intermediate monocytes", "Macrophages", "Megakaryocyte", "Erythroid-like and erythroid precursor cells", "Myeloid Dendritic cells", "Plasmacytoid Dendritic cells", "Granulocytes", "ISG expressing immune cells"))
obj_M72_Myeloid <- subset(obj_M72, subset = sctype_classification %in% c("Eosinophils", "Neutrophils", "Basophils", "Mast cells", "Classical Monocytes", "Non-classical monocytes", "Intermediate monocytes", "Macrophages", "Megakaryocyte", "Erythroid-like and erythroid precursor cells", "Myeloid Dendritic cells", "Plasmacytoid Dendritic cells", "Granulocytes", "ISG expressing immune cells"))
obj_UNSTIM_Myeloid <- subset(obj_UNSTIM, subset = sctype_classification %in% c("Eosinophils", "Neutrophils", "Basophils", "Mast cells", "Classical Monocytes", "Non-classical monocytes", "Intermediate monocytes", "Macrophages", "Megakaryocyte", "Erythroid-like and erythroid precursor cells", "Myeloid Dendritic cells", "Plasmacytoid Dendritic cells", "Granulocytes", "ISG expressing immune cells"))
obj_Myeloid <- subset(obj, subset = sctype_classification %in% c("Eosinophils", "Neutrophils", "Basophils", "Mast cells", "Classical Monocytes", "Non-classical monocytes", "Intermediate monocytes", "Macrophages", "Megakaryocyte", "Erythroid-like and erythroid precursor cells", "Myeloid Dendritic cells", "Plasmacytoid Dendritic cells", "Granulocytes", "ISG expressing immune cells"))

obj_D0_T <- subset(obj_D0, subset = sctype_classification %in% c("Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_D37_T <- subset(obj_D37, subset = sctype_classification %in% c("Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_M72_T <- subset(obj_M72, subset = sctype_classification %in% c("Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_UNSTIM_T <- subset(obj_UNSTIM, subset = sctype_classification %in% c("Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols
obj_T <- subset(obj, subset = sctype_classification %in% c("Naive CD8+ T cells", "Naive CD4+ T cells", "Memory CD8+ T cells", "Memory CD4+ T cells", "Effector CD8+ T cells", "Effector CD4+ T cells", "CD8+ NKT-like cells", "CD4+ NKT-like cells", "Natural killer cells", "Antigen-specific CD8+ T cells", "Antigen-specific CD4+ T cells", "Th1", "Th2", "Th17", "Th1/Th17", "Th22", "Th1 cytotoxic")) #add in gd-T cells but need to know how to put symbols

obj_interest <- subset(obj_M72, subset = sctype_classification %in% c("Naive CD4+ T cells")) # "Classical Monocytes", "Th1", "Effector CD4+ T cells"  
obj_interest <- obj_M72_Lymphoid
  
```

### DEG

```{r}
# Set these as identities in the Seurat obj_interest
obj_interest <- SetIdent(obj_interest, value = obj_interest$VISIT)

# Perform differential expression
de_results <- FindMarkers(
  obj_interest, 
  ident.1 = "D37", 
  ident.2 = "D0", 
  #min.pct = 0.1,   # Min pct of cells where the gene must be detected (adjust as needed)
  logfc.threshold = 0.25,  # Log fold change threshold (adjust as needed)
  test.use = "LR",  # Using logistic regression
  p.value = 0.05
)

# Adjust p-values for multiple testing using Benjamini-Hochberg method
de_results$padj <- p.adjust(de_results$p_val, method = "BH")

# Filter results based on adjusted p-value
significant_genes <- subset(de_results, padj < 0.05)

# Add gene names from row names to a new column called 'gene'
significant_genes$gene <- rownames(significant_genes)

# Sort by p-value and select the top 20
top_genes <- significant_genes %>%
  arrange(p_val) %>%
  slice(1:20)

# Create a volcano plot
ggplot(significant_genes, aes(x=avg_log2FC, y=-log10(p_val))) +
  geom_point(aes(col=padj < 0.1), alpha=0.5) +  # Color points by adjusted p-value threshold
  scale_color_manual(values = c("grey", "red")) +
  theme_minimal() +
  labs(x="Log2 Fold Change", y="-Log10 P-value", title="Volcano Plot of Differential Expression") +
  theme(legend.position="none")

# Create volcano plot with top 20 DEG highlighted in red
# Create the base volcano plot
volcano_plot <- ggplot(significant_genes, aes(x=avg_log2FC, y=-log10(p_val))) +
  geom_point(aes(col=padj < 0.1), alpha=0.5) +  # Color points by adjusted p-value threshold
  scale_color_manual(values = c("grey", "red")) +  # Color non-significant in grey, significant in red
  theme_minimal() +
  labs(x="Log2 Fold Change", y="-Log10 P-value", title="Volcano Plot of Differential Expression") +
  theme(legend.position="none")

# Highlight and label the top 20 genes
volcano_plot <- volcano_plot +
  geom_point(data=top_genes, aes(x=avg_log2FC, y=-log10(p_val)), color="red", size=3) +
  geom_text(data=top_genes, aes(x=avg_log2FC, y=-log10(p_val), label=gene), vjust=1.5, color="black", size=3)

ggsave(filename = paste0(dir_save, "ClusterProfiler_volcanoDEG_EffectorCD4T.png"), plot = volcano_plot, width = 8, height = 5)


# Create heatmaps
# Extract the expression data for significant genes
all_genes <- rownames(obj_interest)
sig_genes <- intersect(all_genes, rownames(significant_genes))
data_matrix <- GetAssayData(obj_interest, layer = "data")[sig_genes, ]

if (!is.matrix(data_matrix)) {
    data_matrix <- as.matrix(data_matrix)
}

# Create the heatmap
#Heatmap(data_matrix, 
        #cluster_rows = TRUE, 
        #cluster_columns = TRUE,
        #show_row_names = TRUE,
        #show_column_names = TRUE,
        #row_title = "Significant Genes", 
        #column_title = "Samples",
        #heatmap_legend_param = list(title = "Expression Level", at = c(min(data_matrix), max(data_matrix)), labels = c("Low", "High")))

# Create heatmap of only top 20 DEG
# Sort significant_genes by adjusted p-value and select the top 20
top_20_genes <- significant_genes[order(significant_genes$padj), ][1:20, ]

# Ensure gene names are in a column if they are currently row names
if("gene" %in% colnames(top_20_genes)) {
    top_20_gene_names <- top_20_genes$gene
} else {
    top_20_gene_names <- rownames(top_20_genes)
}

# Extract the expression data for these top 50 genes
data_matrix <- GetAssayData(obj_interest, slot = "data")[top_20_gene_names, ]

# Convert to matrix if not already
if (!is.matrix(data_matrix)) {
    data_matrix <- as.matrix(data_matrix)
}

# Create the heatmap
#Heatmap(data_matrix,
        #cluster_rows = TRUE,
        #cluster_columns = TRUE,
        #show_row_names = TRUE,
        #show_column_names = TRUE,
        #row_title = "Top 20 Significant Genes",
        #column_title = "Samples",
        #heatmap_legend_param = list(title = "Expression Level", at = c(min(data_matrix, na.rm = TRUE), max(data_matrix, na.rm = TRUE)), labels = c("Low", "High")))



# Dot plots
# Sort by p-value and select the top 20
top_genes <- significant_genes %>%
  arrange(p_val) %>%
  slice(1:20)

# Create a dot plot
dotstats_plot <- ggplot(top_genes, 
               aes(x = gene, y = avg_log2FC, size = -log10(p_val), color = avg_log2FC)) +
  geom_point(alpha=0.6) +
  scale_size(range = c(3, 8)) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_bw() +
  labs(title = "Top 20 Differentially Expressed Genes",
       x = "Gene", y = "Average Log Fold Change") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(filename = paste0(dir_save, "ClusterProfiler_dotDEGstats_EffectorCD4T.png"), plot = dotstats_plot, width = 8, height = 5)


# Create a DotPlot in Seurat
# Extract the 'gene' column as a vector
top_20DEG <- top_genes$gene

dotDEG_plot <- DotPlot(obj_interest, features = top_20DEG, cols = c("blue", "red")) +
  theme_bw() +
  ggtitle("Top 20 Differentially Expressed Genes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

ggsave(filename = paste0(dir_save, "ClusterProfiler_dotDEG_EffectorCD4T.png"), plot = dotDEG_plot, width = 6, height = 4)

# Violin plot for a few selected genes
vln_plot <- VlnPlot(obj_interest, features = c(top_20DEG), group.by = "VISIT",
        pt.size = 1)

ggsave(filename = paste0(dir_save, "ClusterProfiler_vlnDEG_EffectorCD4T.png"), plot = vln_plot, width = 6, height = 8)

```

### Plot DEGs

```{r, echo = FALSE}
#| fig-height: 7
print(volcano_plot)
```

```{r, echo = FALSE}
#| fig-height: 7
print(dotstats_plot)
```

```{r, echo = FALSE}
#| fig-height: 7
print(dotDEG_plot)
```

```{r, echo = FALSE}
#| fig-height: 21
print(vln_plot)
```

### clusterProfiler

```{r, message = FALSE, warning = FALSE, output = FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(DOSE)
library(ggplot2)
library(enrichplot)

# Prepare gene list for GSEA (ranks based on logFC or other metric)
gene_list <- de_results$avg_log2FC
names(gene_list) <- rownames(de_results)
gene_list <- sort(gene_list, decreasing = TRUE)

# Convert gene symbols to Entrez IDs for clusterProfiler
#top_genes_list <- top_genes$gene
#top_genes_list_entrez <- bitr(top_genes_list, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db, drop = TRUE)
sig_genes <- significant_genes$gene
sig_genes_entrez <- bitr(sig_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db, drop = TRUE)
gene_list_entrez <- bitr(names(gene_list), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db, drop = TRUE)
gene_list <- setNames(gene_list, gene_list_entrez$ENTREZID)

# Remove any NA values from gene_list
gene_list <- gene_list[!is.na(names(gene_list))]

# Verify gene_list content
head(gene_list)

# Over-representation analysis (ORA)
ego <- enrichGO(gene = sig_genes_entrez$ENTREZID,
                OrgDb = org.Hs.eg.db,
                ont = "BP", # Biological Process
                pAdjustMethod = "BH",
                pvalueCutoff = 0.05,
                qvalueCutoff = 0.05,
                readable = TRUE)


# Plot ORA results
ora_dot_plot <- dotplot(ego, showCategory = 20) + ggtitle("Over-representation Analysis")
ora_bar_plot <- barplot(ego, showCategory=20) 

# Enrichment map
# Compute the term similarity matrix
ego <- pairwise_termsim(ego)
ora_emap_plot <- emapplot(ego) + ggtitle("Enrichment Map of GO Terms")

# Gene-concept network
## convert gene ID to Symbol
egox <- setReadable(ego, 'org.Hs.eg.db', 'ENTREZID')
p1 <- cnetplot(egox, color.params = list(foldChange = gene_list))
## categorySize can be scaled by 'pvalue' or 'geneNum'
p2 <- cnetplot(egox, categorySize="pvalue", color.params = list(foldChange = gene_list)) + 
    scale_color_gradient2(name='fold change', low='steelblue', high='firebrick')
p3 <- cnetplot(egox, color.params = list(foldChange = gene_list), circular = TRUE, colorEdge = TRUE) + 
    scale_color_gradient2(name='fold change', low='steelblue', high='firebrick')
ora_gcn_plot <- cowplot::plot_grid(p2, p3, ncol=2, labels=LETTERS[1:2], rel_widths=c(.8, 1.2))
ggsave(filename = paste0(dir_save, "ClusterProfiler_ORA_cnetplots.png"), plot = ora_gcn_plot, width = 25, height = 12)


# Heatmap classification
p1 <- heatplot(egox, showCategory=5)
p2 <- heatplot(egox, foldChange=gene_list, showCategory=10) +
    scale_fill_gradient2(name='fold change', low='steelblue', high='firebrick')
ora_hm_plot <- cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
ggsave(filename = paste0(dir_save, "ClusterProfiler_ORA_heatmaps.png"), plot = ora_hm_plot, width = 25, height = 10)


# Tree plot
egox2 <- pairwise_termsim(egox)
p1 <- treeplot(egox2)
p2 <- treeplot(egox2, hclust_method = "average")
ora_tree_plot <- aplot::plot_list(p1, p2, tag_levels='A')
ggsave(filename = paste0(dir_save, "ClusterProfiler_ORA_treeplots.png"), plot = ora_tree_plot, width = 25, height = 10)

# Upset plots
#upsetplot(ego)


# Gene Set Enrichment Analysis (GSEA)
gsea_res <- gseGO(geneList = gene_list,
                  OrgDb = org.Hs.eg.db,
                  ont = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.05,
                  verbose = FALSE,
                  by = "fgsea")

# Save GSEA result to an RDS file
saveRDS(gsea_res, file = paste0(dir_save, "gsea_results.rds"))

# Plot GSEA results
gsea_dot_plot <- dotplot(gsea_res, showCategory=30) + ggtitle("dotplot for GSEA")
ggsave(filename = paste0(dir_save, "ClusterProfiler_GSEA_dotplot.png"), plot = gsea_dot_plot, width = 8, height = 12)

# Gene-concept network
## convert gene ID to Symbol
egox <- setReadable(gsea_res, 'org.Hs.eg.db', 'ENTREZID')
p1 <- cnetplot(egox, color.params = list(foldChange = gene_list))
## categorySize can be scaled by 'pvalue' or 'geneNum'
p2 <- cnetplot(egox, categorySize="pvalue", color.params = list(foldChange = gene_list)) + 
    scale_color_gradient2(name='fold change', low='steelblue', high='firebrick')
p3 <- cnetplot(egox, color.params = list(foldChange = gene_list), circular = TRUE, colorEdge = TRUE) + 
    scale_color_gradient2(name='fold change', low='steelblue', high='firebrick')
gsea_gcn_plot <- cowplot::plot_grid(p2, p3, ncol=2, labels=LETTERS[1:2], rel_widths=c(.8, 1.2))
ggsave(filename = paste0(dir_save, "ClusterProfiler_GSEA_cnetplots.png"), plot = gsea_gcn_plot, width = 25, height = 12)

# Heatmap classification
p1 <- heatplot(egox, showCategory=5)
p2 <- heatplot(egox, foldChange=gene_list, showCategory=5) +
    scale_fill_gradient2(name='fold change', low='steelblue', high='firebrick')
gsea_hm_plot <- cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
ggsave(filename = paste0(dir_save, "ClusterProfiler_GSEA_heatmaps.png"), plot = gsea_hm_plot, width = 25, height = 10)

# Ridgeplot
gsea_ridge_plot <- ridgeplot(gsea_res)
ggsave(filename = paste0(dir_save, "ClusterProfiler_GSEA_ridgeplot.png"), plot = gsea_ridge_plot, width = 10, height = 15)

# Running score and preranked list
p1 <- gseaplot(gsea_res, geneSetID = 1, by = "runningScore", title = gsea_res$Description[1])
p2 <- gseaplot(gsea_res, geneSetID = 1, by = "preranked", title = gsea_res$Description[1])
p3 <- gseaplot(gsea_res, geneSetID = 1, title = gsea_res$Description[1])
cowplot::plot_grid(p1, p2, p3, ncol=1, labels=LETTERS[1:3])

gseaplot2(gsea_res, geneSetID = 1, title = gsea_res$Description[1])
gseaplot2(gsea_res, geneSetID = 1:3)
gsea_rs_plot <- gseaplot2(gsea_res, geneSetID = 1:3, pvalue_table = TRUE,
          color = c("#E495A5", "#86B875", "#7DB0DD"), ES_geom = "dot")
ggsave(filename = paste0(dir_save, "ClusterProfiler_GSEA_gseaplot.png"), plot = gsea_rs_plot, width = 15, height = 15)


```

### Plot ORA results

```{r, echo = FALSE}
#| fig-height: 10
print(ora_dot_plot)
```

```{r, echo = FALSE}
#| fig-height: 10
print(ora_bar_plot)
```

```{r, echo = FALSE}
#| fig-height: 10
print(ora_emap_plot)
```

```{r, echo = FALSE, warning = FALSE}
#| fig-height: 10
print(ora_gcn_plot)
```

```{r, echo = FALSE}
#| fig-height: 7
print(ora_hm_plot)
```

```{r, echo = FALSE}
#| fig-height: 5
print(ora_tree_plot)
```

### Plot GSEA results

```{r, echo = FALSE}
#| fig-height: 10
print(gsea_dot_plot)
```

```{r, echo = FALSE, warning = FALSE}
#| fig-height: 10
print(gsea_gcn_plot)
```

```{r, echo = FALSE}
#| fig-height: 7
print(gsea_hm_plot)
```

```{r, echo = FALSE}
#| fig-height: 10
print(gsea_ridge_plot)
```

```{r, echo = FALSE}
#| fig-height: 10
print(gsea_rs_plot)
```

## DEG and clusterProfiler for all cell types and STIMs

```{r}
# Identify unique features in obj@meta.data$sctype_classification and obj@meta.data$STIM
  sctype_classifications <- setdiff(unique(obj$sctype_classification), c("Basophils", "Plasma B cells"))
  #sctype_classifications <- c("Effector CD4+ T cells")
  stim_conditions <- setdiff(unique(obj$STIM), "UNSTIM")
  
  # Directory to save plots
  dir_save <- "output/dataset_full/"
  
  # Outer loop through each STIM condition
  for (stim in stim_conditions) {
    # Subset the Seurat obj based on STIM condition
    obj_stim <- subset(obj, STIM == stim)
    
    # Inner loop through each cell type in sctype_classification
    for (cell_type in sctype_classifications) {
      # Subset the Seurat obj based on sctype_classification
      obj_subset <- subset(obj_stim, sctype_classification == cell_type)
      
      # Set these as identities in the Seurat obj_subset
      obj_subset <- SetIdent(obj_subset, value = obj_subset$VISIT)
      
      # Perform differential expression
      de_results <- FindMarkers(
        obj_subset, 
        ident.1 = "D37", 
        ident.2 = "D0", 
        logfc.threshold = 0.25,  # Log fold change threshold (adjust as needed)
        test.use = "LR",  # Using logistic regression
        p.value = 0.05
      )
      
      # Adjust p-values for multiple testing using Benjamini-Hochberg method
      de_results$padj <- p.adjust(de_results$p_val, method = "BH")
      
      # Filter results based on adjusted p-value
      significant_genes <- subset(de_results, padj < 0.05)
      
      # Add gene names from row names to a new column called 'gene'
      significant_genes$gene <- rownames(significant_genes)
      
      # Determine the number of top genes to use
      num_top_genes <- min(20, nrow(significant_genes))
      
      # Sort by p-value and select the top genes
      top_genes <- significant_genes %>%
        arrange(p_val) %>%
        slice(1:num_top_genes)
      
      # Create a volcano plot
      volcano_plot <- ggplot(significant_genes, aes(x=avg_log2FC, y=-log10(p_val))) +
        geom_point(aes(col=padj < 0.1), alpha=0.5) +  # Color points by adjusted p-value threshold
        scale_color_manual(values = c("grey", "red")) +
        theme_minimal() +
        labs(x="Log2 Fold Change", y="-Log10 P-value", title=paste("Volcano Plot of Differential Expression for", cell_type, "under", stim, "condition")) +
        theme(legend.position="none")
      
      # Highlight and label the top genes
      volcano_plot <- volcano_plot +
        geom_point(data=top_genes, aes(x=avg_log2FC, y=-log10(p_val)), color="red", size=3) +
        geom_text(data=top_genes, aes(x=avg_log2FC, y=-log10(p_val), label=gene), vjust=1.5, color="black", size=3)
      
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_volcanoDEG.png"), plot = volcano_plot, width = 8, height = 10)
      
      # Create heatmaps
      # Extract the expression data for significant genes
      all_genes <- rownames(obj_subset)
      sig_genes <- intersect(all_genes, rownames(significant_genes))
      data_matrix <- GetAssayData(obj_subset, slot = "data")[sig_genes, ]
      
      if (!is.matrix(data_matrix)) {
        data_matrix <- as.matrix(data_matrix)
      }
      
      # Create the heatmap
      #Heatmap(data_matrix, 
      #       cluster_rows = TRUE, 
      #      cluster_columns = TRUE,
      #     show_row_names = TRUE,
      #   show_column_names = TRUE,
      #   row_title = "Significant Genes", 
      #  column_title = "Samples",
      # heatmap_legend_param = list(title = "Expression Level", at = c(min(data_matrix, na.rm = TRUE), max(data_matrix, na.rm = TRUE)), labels = c("Low", "High")))
      
      # Create heatmap of only top genes
      # Ensure gene names are in a column if they are currently row names
      if("gene" %in% colnames(top_genes)) {
        top_gene_names <- top_genes$gene
      } else {
        top_gene_names <- rownames(top_genes)
      }
      
      # Extract the expression data for these top genes
      data_matrix <- GetAssayData(obj_subset, slot = "data")[top_gene_names, ]
      
      # Convert to matrix if not already
      if (!is.matrix(data_matrix)) {
        data_matrix <- as.matrix(data_matrix)
      }
      
      # Create the heatmap
      #Heatmap(data_matrix,
      #       cluster_rows = TRUE,
      #      cluster_columns = TRUE,
      #     show_row_names = TRUE,
      #    show_column_names = TRUE,
      #   row_title = "Top Significant Genes",
      #  column_title = "Samples",
      # heatmap_legend_param = list(title = "Expression Level", at = c(min(data_matrix, na.rm = TRUE), max(data_matrix, na.rm = TRUE)), labels = c("Low", "High")))
      
      # Dot plots
      # Sort by p-value and select the top genes
      top_genes <- significant_genes %>%
        arrange(p_val) %>%
        slice(1:num_top_genes)
      
      # Create a dot plot
      plot <- ggplot(top_genes, 
                     aes(x = gene, y = avg_log2FC, size = -log10(p_val), color = avg_log2FC)) +
        geom_point(alpha=0.6) +
        scale_size(range = c(3, 8)) +
        scale_color_gradient(low = "blue", high = "red") +
        theme_bw() +
        labs(title = paste("Top Differentially Expressed Genes for", cell_type, "under", stim, "condition"),
             x = "Gene", y = "Average Log Fold Change") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
      
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_dotDEGstats.png"), plot = plot, width = 8, height = 5)
      
      # Create a DotPlot in Seurat
      # Extract the 'gene' column as a vector
      top_DEG <- top_genes$gene
      
      # Check if top_DEG contains any genes before plotting
      if (length(top_DEG) > 0) {
        plot <- DotPlot(obj_subset, features = top_DEG, cols = c("blue", "red")) +
          theme_bw() +
          ggtitle(paste("Top Differentially Expressed Genes for", cell_type, "under", stim, "condition")) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
                axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
                axis.title.x = element_blank(),
                axis.title.y = element_blank())
    
        # Extract the legends
        legend_color <- cowplot::get_legend(plot + guides(size = "none"))
        legend_size <- cowplot::get_legend(plot + guides(color = "none"))
    
        # Combine the legends with some vertical space
        legends <- cowplot::plot_grid(legend_color, legend_size, ncol = 1, rel_heights = c(1, 1), align = "v")
    
        # Combine the plot and legends
        combined_plot <- cowplot::plot_grid(
          plot + theme(legend.position = "none"),
          cowplot::plot_grid(NULL, legends, ncol = 1, rel_heights = c(0.03, 1)),
          ncol = 2,
          rel_widths = c(3, 0.75)
        )
    
        ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_dotDEG.png"), plot = combined_plot, width = 9, height = 3)
      } else {
          print(paste("No top DEGs found for DotPlot for", cell_type, "under", stim, "condition"))
      }       
      
      # Violin plot for a few selected genes
      plot <- VlnPlot(obj_subset, features = c(top_DEG), group.by = "VISIT",
                      pt.size = 0)
      
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_violinDEG.png"), plot = plot, width = 8, height = 30)
      
      # ClusterProfiler Analysis
      # Prepare gene list for GSEA (ranks based on logFC or other metric)
      gene_list <- de_results$avg_log2FC
      names(gene_list) <- rownames(de_results)
      gene_list <- sort(gene_list, decreasing = TRUE)
      
      # Convert gene symbols to Entrez IDs for clusterProfiler
      #top_genes_list <- top_genes$gene
      #top_genes_list_entrez <- bitr(top_genes_list, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db, drop = TRUE)
      sig_genes <- significant_genes$gene
      sig_genes_entrez <- bitr(sig_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
      gene_list_entrez <- bitr(names(gene_list), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
      gene_list <- setNames(gene_list, gene_list_entrez$ENTREZID)
      
      # Remove any NA values from gene_list
      gene_list <- gene_list[!is.na(names(gene_list))]
      
      # Verify gene_list content
      head(gene_list)
      
      # Over-representation analysis (ORA)
      ego <- enrichGO(gene = sig_genes_entrez$ENTREZID,
                      OrgDb = org.Hs.eg.db,
                      ont = "BP", # Biological Process
                      pAdjustMethod = "BH",
                      pvalueCutoff = 0.05,
                      qvalueCutoff = 0.05,
                      readable = TRUE)
      
      # Check if ego contains any enriched terms before plotting
      if (nrow(ego) > 0) {
        
        # Plot ORA results
        ora_dot_plot <- dotplot(ego, showCategory = 20) + ggtitle("Over-representation Analysis")
        ora_bar_plot <- barplot(ego, showCategory=20) 
        
        # Enrichment map
        # Compute the term similarity matrix
        ego <- pairwise_termsim(ego)
        ora_emap_plot <- emapplot(ego) + ggtitle("Enrichment Map of GO Terms")
        
        # Gene-concept network
        ## convert gene ID to Symbol
        egox <- setReadable(ego, 'org.Hs.eg.db', 'ENTREZID')
        p1 <- cnetplot(egox, color.params = list(foldChange = gene_list))
        ## categorySize can be scaled by 'pvalue' or 'geneNum'
        p2 <- cnetplot(egox, categorySize="pvalue", color.params = list(foldChange = gene_list)) + 
          scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
        p3 <- cnetplot(egox, color.params = list(foldChange = gene_list), circular = TRUE, colorEdge = TRUE) + 
          scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
        ora_gcn_plot <- cowplot::plot_grid(p2, p3, ncol=2, labels=LETTERS[1:2], rel_widths=c(.8, 1.2))
        ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_ORA_cnetplots.png"), plot = ora_gcn_plot, width = 25, height = 12)
        
        # Heatmap classification
        p1 <- heatplot(egox, showCategory=5)
        p2 <- heatplot(egox, foldChange=gene_list, showCategory=10) +
          scale_fill_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
        ora_hm_plot <- cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
        ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_ORA_heatmaps.png"), plot = ora_hm_plot, width = 25, height = 10)
        
        # Tree plot
        #egox2 <- pairwise_termsim(egox)
        #if (nrow(egox2) > 3) {
          #p1 <- treeplot(egox2)
          #p2 <- treeplot(egox2, hclust_method = "average")
          #plot <- aplot::plot_list(p1, p2, tag_levels='A')
          #ggsave(filename = paste0(dir_save, cell_type, "_", stim, "_ORA_treeplots.png"), plot = plot, width = 25, height = 10)
        #} else {
          #print(paste("Skipping tree plot for", cell_type, "under", stim, "condition due to insufficient clusters"))
        #}
      #} else {
        #print(paste("No enriched terms found for", cell_type, "under", stim, "condition"))
      #}
      
      # Gene Set Enrichment Analysis (GSEA)
      gsea_res <- gseGO(geneList = gene_list,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        pvalueCutoff = 1, #CHANGE THIS!!!
                        verbose = FALSE,
                        by = "fgsea")
      
      # Save GSEA result to an RDS file
      saveRDS(gsea_res, file = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_gsea_results.rds"))
      
      # Plot GSEA results
      gsea_dot_plot <- dotplot(gsea_res, showCategory=30) + ggtitle("dotplot for GSEA")
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_GSEA_dotplot.png"), plot = gsea_dot_plot, width = 8, height = 12)
      
      # Gene-concept network
      ## convert gene ID to Symbol
      egox <- setReadable(gsea_res, 'org.Hs.eg.db', 'ENTREZID')
      p1 <- cnetplot(egox, color.params = list(foldChange = gene_list))
      ## categorySize can be scaled by 'pvalue' or 'geneNum'
      p2 <- cnetplot(egox, categorySize="pvalue", color.params = list(foldChange = gene_list)) + 
        scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
      p3 <- cnetplot(egox, color.params = list(foldChange = gene_list), circular = TRUE, colorEdge = TRUE) + 
        scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
      gsea_gcn_plot <- cowplot::plot_grid(p2, p3, ncol=2, labels=LETTERS[1:2], rel_widths=c(.8, 1.2))
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_GSEA_cnetplots.png"), plot = gsea_gcn_plot, width = 25, height = 12)
      
      # Heatmap classification
      p1 <- heatplot(egox, showCategory=5)
      p2 <- heatplot(egox, foldChange=gene_list, showCategory=5) +
        scale_fill_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
      gsea_hm_plot <- cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_GSEA_heatmaps.png"), plot = gsea_hm_plot, width = 25, height = 10)
      
      # Ridgeplot
      gsea_ridge_plot <- ridgeplot(gsea_res)
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_GSEA_ridgeplot.png"), plot = gsea_ridge_plot, width = 10, height = 20)
      
      # Running score and preranked list
      p1 <- gseaplot(gsea_res, geneSetID = 1, by = "runningScore", title = gsea_res$Description[1])
      p2 <- gseaplot(gsea_res, geneSetID = 1, by = "preranked", title = gsea_res$Description[1])
      p3 <- gseaplot(gsea_res, geneSetID = 1, title = gsea_res$Description[1])
      cowplot::plot_grid(p1, p2, p3, ncol=1, labels=LETTERS[1:3])
      
      gseaplot2(gsea_res, geneSetID = 1, title = gsea_res$Description[1])
      gseaplot2(gsea_res, geneSetID = 1:3)
      gsea_rs_plot <- gseaplot2(gsea_res, geneSetID = 1:3, pvalue_table = TRUE,
                        color = c("#E495A5", "#86B875", "#7DB0DD"), ES_geom = "dot")
      ggsave(filename = paste0(dir_save, "ClusterProfiler_", cell_type, "_", stim, "_GSEA_gseaplot.png"), plot = gsea_rs_plot, width = 15, height = 15)
    }
    }
  }
```













## DEG and clusterProfiler for all clusters and STIMs

```{r}
# Identify unique features in obj@meta.data$feature_based_ID_letter and obj@STIM
  #clusters <- unique(obj$feature_based_ID_letter)
  clusters<- setdiff(unique(obj$feature_based_ID_letter), c("S", "L", "R"))
  stim_conditions <- setdiff(unique(obj$STIM), "UNSTIM")
  
  # Directory to save plots
  dir_save <- "output/dataset_full/"
  
  # Outer loop through each STIM condition
  for (stim in stim_conditions) {
    # Subset the Seurat obj based on STIM condition
    obj_stim <- subset(obj, STIM == stim)
    
    # Inner loop through each cluster in clusters
    for (cluster in clusters) {
      # Subset the Seurat obj based on clusters
      obj_subset <- subset(obj_stim, feature_based_ID_letter == cluster)
      
      # Set these as identities in the Seurat obj_subset
      obj_subset <- SetIdent(obj_subset, value = obj_subset$VISIT)
      
      # Perform differential expression
      de_results <- FindMarkers(
        obj_subset, 
        ident.1 = "D37", 
        ident.2 = "D0", 
        logfc.threshold = 0.25,  # Log fold change threshold (adjust as needed)
        test.use = "LR",  # Using logistic regression
        p.value = 0.05
      )
      
      # Adjust p-values for multiple testing using Benjamini-Hochberg method
      de_results$padj <- p.adjust(de_results$p_val, method = "BH")
      
      # Filter results based on adjusted p-value
      significant_genes <- subset(de_results, padj < 0.05)
      
      # Add gene names from row names to a new column called 'gene'
      significant_genes$gene <- rownames(significant_genes)
      
      # Determine the number of top genes to use
      num_top_genes <- min(20, nrow(significant_genes))
      
      # Sort by p-value and select the top genes
      top_genes <- significant_genes %>%
        arrange(p_val) %>%
        slice(1:num_top_genes)
      
      # Create a volcano plot
      volcano_plot <- ggplot(significant_genes, aes(x=avg_log2FC, y=-log10(p_val))) +
        geom_point(aes(col=padj < 0.1), alpha=0.5) +  # Color points by adjusted p-value threshold
        scale_color_manual(values = c("grey", "red")) +
        theme_minimal() +
        labs(x="Log2 Fold Change", y="-Log10 P-value", title=paste("Volcano Plot of Differential Expression for cluster", cluster, "under", stim, "condition")) +
        theme(legend.position="none")
      
      # Highlight and label the top genes
      volcano_plot <- volcano_plot +
        geom_point(data=top_genes, aes(x=avg_log2FC, y=-log10(p_val)), color="red", size=3) +
        geom_text(data=top_genes, aes(x=avg_log2FC, y=-log10(p_val), label=gene), vjust=1.5, color="black", size=3)
      
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_volcanoDEG.png"), plot = volcano_plot, width = 8, height = 10)
      
      # Create heatmaps
      # Extract the expression data for significant genes
      all_genes <- rownames(obj_subset)
      sig_genes <- intersect(all_genes, rownames(significant_genes))
      data_matrix <- GetAssayData(obj_subset, slot = "data")[sig_genes, ]
      
      if (!is.matrix(data_matrix)) {
        data_matrix <- as.matrix(data_matrix)
      }
      
      # Create the heatmap
      #Heatmap(data_matrix, 
      #       cluster_rows = TRUE, 
      #      cluster_columns = TRUE,
      #     show_row_names = TRUE,
      #   show_column_names = TRUE,
      #   row_title = "Significant Genes", 
      #  column_title = "Samples",
      # heatmap_legend_param = list(title = "Expression Level", at = c(min(data_matrix, na.rm = TRUE), max(data_matrix, na.rm = TRUE)), labels = c("Low", "High")))
      
      # Create heatmap of only top genes
      # Ensure gene names are in a column if they are currently row names
      if("gene" %in% colnames(top_genes)) {
        top_gene_names <- top_genes$gene
      } else {
        top_gene_names <- rownames(top_genes)
      }
      
      # Extract the expression data for these top genes
      data_matrix <- GetAssayData(obj_subset, slot = "data")[top_gene_names, ]
      
      # Convert to matrix if not already
      if (!is.matrix(data_matrix)) {
        data_matrix <- as.matrix(data_matrix)
      }
      
      # Create the heatmap
      #Heatmap(data_matrix,
      #       cluster_rows = TRUE,
      #      cluster_columns = TRUE,
      #     show_row_names = TRUE,
      #    show_column_names = TRUE,
      #   row_title = "Top Significant Genes",
      #  column_title = "Samples",
      # heatmap_legend_param = list(title = "Expression Level", at = c(min(data_matrix, na.rm = TRUE), max(data_matrix, na.rm = TRUE)), labels = c("Low", "High")))
      
      # Dot plots
      # Sort by p-value and select the top genes
      top_genes <- significant_genes %>%
        arrange(p_val) %>%
        slice(1:num_top_genes)
      
      # Create a dot plot
      plot <- ggplot(top_genes, 
                     aes(x = gene, y = avg_log2FC, size = -log10(p_val), color = avg_log2FC)) +
        geom_point(alpha=0.6) +
        scale_size(range = c(3, 8)) +
        scale_color_gradient(low = "blue", high = "red") +
        theme_bw() +
        labs(title = paste("Top Differentially Expressed Genes for cluster", cluster, "under", stim, "condition"),
             x = "Gene", y = "Average Log Fold Change") +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
      
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_dotDEGstats.png"), plot = plot, width = 8, height = 5)
      
      # Create a DotPlot in Seurat
      # Extract the 'gene' column as a vector
      top_DEG <- top_genes$gene
      
      # Check if top_DEG contains any genes before plotting
      if (length(top_DEG) > 0) {
        plot <- DotPlot(obj_subset, features = top_DEG, cols = c("blue", "red")) +
          theme_bw() +
          ggtitle(paste("Top Differentially Expressed Genes for cluster", cluster, "under", stim, "condition")) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
                axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
                axis.title.x = element_blank(),
                axis.title.y = element_blank())
    
        # Extract the legends
        legend_color <- cowplot::get_legend(plot + guides(size = "none"))
        legend_size <- cowplot::get_legend(plot + guides(color = "none"))
    
        # Combine the legends with some vertical space
        legends <- cowplot::plot_grid(legend_color, legend_size, ncol = 1, rel_heights = c(1, 1), align = "v")
    
        # Combine the plot and legends
        combined_plot <- cowplot::plot_grid(
          plot + theme(legend.position = "none"),
          cowplot::plot_grid(NULL, legends, ncol = 1, rel_heights = c(0.03, 1)),
          ncol = 2,
          rel_widths = c(3, 0.75)
        )
    
        ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_dotDEG.png"), plot = combined_plot, width = 9, height = 3)
      } else {
          print(paste("No top DEGs found for DotPlot for cluster", cluster, "under", stim, "condition"))
      }       
      
      # Violin plot for a few selected genes
      plot <- VlnPlot(obj_subset, features = c(top_DEG), group.by = "VISIT",
                      pt.size = 0)
      
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_violinDEG.png"), plot = plot, width = 8, height = 30)
      
      # ClusterProfiler Analysis
      # Prepare gene list for GSEA (ranks based on logFC or other metric)
      gene_list <- de_results$avg_log2FC
      names(gene_list) <- rownames(de_results)
      gene_list <- sort(gene_list, decreasing = TRUE)
      
      # Convert gene symbols to Entrez IDs for clusterProfiler
      #top_genes_list <- top_genes$gene
      #top_genes_list_entrez <- bitr(top_genes_list, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db, drop = TRUE)
      sig_genes <- significant_genes$gene
      sig_genes_entrez <- bitr(sig_genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
      gene_list_entrez <- bitr(names(gene_list), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
      gene_list <- setNames(gene_list, gene_list_entrez$ENTREZID)
      
      # Remove any NA values from gene_list
      gene_list <- gene_list[!is.na(names(gene_list))]
      
      # Verify gene_list content
      head(gene_list)
      
      # Over-representation analysis (ORA)
      ego <- enrichGO(gene = sig_genes_entrez$ENTREZID,
                      OrgDb = org.Hs.eg.db,
                      ont = "BP", # Biological Process
                      pAdjustMethod = "BH",
                      pvalueCutoff = 0.05,
                      qvalueCutoff = 0.05,
                      readable = TRUE)
      
      # Check if ego contains any enriched terms before plotting
      if (nrow(ego) > 0) {
        
        # Plot ORA results
        ora_dot_plot <- dotplot(ego, showCategory = 20) + ggtitle("Over-representation Analysis")
        ora_bar_plot <- barplot(ego, showCategory=20) 
        
        # Enrichment map
        # Compute the term similarity matrix
        ego <- pairwise_termsim(ego)
        ora_emap_plot <- emapplot(ego) + ggtitle("Enrichment Map of GO Terms")
        
        # Gene-concept network
        ## convert gene ID to Symbol
        egox <- setReadable(ego, 'org.Hs.eg.db', 'ENTREZID')
        p1 <- cnetplot(egox, color.params = list(foldChange = gene_list))
        ## categorySize can be scaled by 'pvalue' or 'geneNum'
        p2 <- cnetplot(egox, categorySize="pvalue", color.params = list(foldChange = gene_list)) + 
          scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
        p3 <- cnetplot(egox, color.params = list(foldChange = gene_list), circular = TRUE, colorEdge = TRUE) + 
          scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
        ora_gcn_plot <- cowplot::plot_grid(p2, p3, ncol=2, labels=LETTERS[1:2], rel_widths=c(.8, 1.2))
        ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_ORA_cnetplots.png"), plot = ora_gcn_plot, width = 25, height = 12)
        
        # Heatmap classification
        p1 <- heatplot(egox, showCategory=5)
        p2 <- heatplot(egox, foldChange=gene_list, showCategory=10) +
          scale_fill_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
        ora_hm_plot <- cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
        ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_ORA_heatmaps.png"), plot = ora_hm_plot, width = 25, height = 10)
        
        # Tree plot
        #egox2 <- pairwise_termsim(egox)
        #if (nrow(egox2) > 3) {
          #p1 <- treeplot(egox2)
          #p2 <- treeplot(egox2, hclust_method = "average")
          #plot <- aplot::plot_list(p1, p2, tag_levels='A')
          #ggsave(filename = paste0(dir_save, cell_type, "_", stim, "_ORA_treeplots.png"), plot = plot, width = 25, height = 10)
        #} else {
          #print(paste("Skipping tree plot for", cell_type, "under", stim, "condition due to insufficient clusters"))
        #}
      #} else {
        #print(paste("No enriched terms found for", cell_type, "under", stim, "condition"))
      #}
      
      # Gene Set Enrichment Analysis (GSEA)
      gsea_res <- gseGO(geneList = gene_list,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        pvalueCutoff = 1, #CHANGE THIS!!!
                        verbose = FALSE,
                        by = "fgsea")
      
      # Save GSEA result to an RDS file
      saveRDS(gsea_res, file = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_gsea_results.rds"))
      
      # Plot GSEA results
      gsea_dot_plot <- dotplot(gsea_res, showCategory=30) + ggtitle("dotplot for GSEA")
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_GSEA_dotplot.png"), plot = gsea_dot_plot, width = 8, height = 12)
      
      # Gene-concept network
      ## convert gene ID to Symbol
      egox <- setReadable(gsea_res, 'org.Hs.eg.db', 'ENTREZID')
      p1 <- cnetplot(egox, color.params = list(foldChange = gene_list))
      ## categorySize can be scaled by 'pvalue' or 'geneNum'
      p2 <- cnetplot(egox, categorySize="pvalue", color.params = list(foldChange = gene_list)) + 
        scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
      p3 <- cnetplot(egox, color.params = list(foldChange = gene_list), circular = TRUE, colorEdge = TRUE) + 
        scale_color_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
      gsea_gcn_plot <- cowplot::plot_grid(p2, p3, ncol=2, labels=LETTERS[1:2], rel_widths=c(.8, 1.2))
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_GSEA_cnetplots.png"), plot = gsea_gcn_plot, width = 25, height = 12)
      
      # Heatmap classification
      p1 <- heatplot(egox, showCategory=5)
      p2 <- heatplot(egox, foldChange=gene_list, showCategory=5) +
        scale_fill_gradient2(name='Log2 fold change', low='steelblue', high='firebrick')
      gsea_hm_plot <- cowplot::plot_grid(p1, p2, ncol=1, labels=LETTERS[1:2])
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_GSEA_heatmaps.png"), plot = gsea_hm_plot, width = 25, height = 10)
      
      # Ridgeplot
      gsea_ridge_plot <- ridgeplot(gsea_res)
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_GSEA_ridgeplot.png"), plot = gsea_ridge_plot, width = 10, height = 20)
      
      # Running score and preranked list
      p1 <- gseaplot(gsea_res, geneSetID = 1, by = "runningScore", title = gsea_res$Description[1])
      p2 <- gseaplot(gsea_res, geneSetID = 1, by = "preranked", title = gsea_res$Description[1])
      p3 <- gseaplot(gsea_res, geneSetID = 1, title = gsea_res$Description[1])
      cowplot::plot_grid(p1, p2, p3, ncol=1, labels=LETTERS[1:3])
      
      gseaplot2(gsea_res, geneSetID = 1, title = gsea_res$Description[1])
      gseaplot2(gsea_res, geneSetID = 1:3)
      gsea_rs_plot <- gseaplot2(gsea_res, geneSetID = 1:3, pvalue_table = TRUE,
                        color = c("#E495A5", "#86B875", "#7DB0DD"), ES_geom = "dot")
      ggsave(filename = paste0(dir_save, "ClusterProfiler_cluster", cluster, "_", stim, "_GSEA_gseaplot.png"), plot = gsea_rs_plot, width = 15, height = 15)
    }
    }
  }
```


### Plot proportion of cell types over feature based clusters

```{r}
# Assign feature_based_ID as the new identities
Idents(obj) <- obj@meta.data$feature_based_ID

# Create a mapping of each feature_based_ID to a letter based on the number of cells
feature_based_ID_counts <- table(Idents(obj))
sorted_feature_based_IDs <- names(sort(feature_based_ID_counts, decreasing = TRUE))
label_mapping <- setNames(LETTERS[1:length(sorted_feature_based_IDs)], sorted_feature_based_IDs)

# Add a new column with the letter mapping
obj@meta.data$feature_based_ID_letter <- label_mapping[as.character(Idents(obj))]

# Verify the change
head(Idents(obj))
head(obj@meta.data$feature_based_ID_letter)

plot_FBID_clusters = function (srat, batchcolumn) {
  ## take an integrated Seurat object, plot distributions over feature_based_ID

  # Ensure the feature_based_ID column is treated as a factor and handle NA values
  srat@meta.data$feature_based_ID_letter <- as.factor(srat@meta.data$feature_based_ID_letter)
  srat@meta.data <- srat@meta.data[!is.na(srat@meta.data$feature_based_ID_letter), ]

  # Ensure the batchcolumn is treated as a factor
  srat@meta.data[[batchcolumn]] <- as.factor(srat@meta.data[[batchcolumn]])

  # Check lengths of the columns
  if (length(srat@meta.data$feature_based_ID_letter) != length(srat@meta.data[[batchcolumn]])) {
    stop("The lengths of feature_based_ID_letter and the batch column do not match.")
  }

  count_table <- table(srat@meta.data$feature_based_ID_letter, srat@meta.data[[batchcolumn]])
  count_mtx   <- as.data.frame.matrix(count_table)
  count_mtx$cluster <- rownames(count_mtx)
  melt_mtx    <- melt(count_mtx)
  melt_mtx$cluster <- as.factor(melt_mtx$cluster)

  # Calculate the total number of cells per cluster and sort by size in descending order
  cluster_size   <- aggregate(value ~ cluster, data = melt_mtx, FUN = sum)
  sorted_labels <- cluster_size$cluster[order(cluster_size$value, decreasing = F)]
  cluster_size$cluster <- factor(cluster_size$cluster, levels = sorted_labels)
  melt_mtx$cluster <- factor(melt_mtx$cluster, levels = sorted_labels)

  colnames(melt_mtx)[2] <- "dataset"

  p1 <- ggplot(cluster_size, aes(y = cluster, x = value)) + 
    geom_bar(position = "dodge", stat = "identity", fill = "grey60") +
    geom_text(aes(label = value), hjust = 1.5, size = 3) +
    theme_bw() + scale_x_log10() + 
    xlab("Cells per cluster, log10 scale") + ylab("")
  p2 <- ggplot(melt_mtx, aes(x = cluster, y = value, fill = dataset)) +
    geom_bar(position = "fill", stat = "identity") + theme_bw() + coord_flip() +
    ylab("Fraction of cells in each dataset") + xlab("Cluster") + theme(legend.position = "left")

  # Create a legend as a text grob using the existing feature_based_ID_letter column
  legend_text <- paste(label_mapping, names(label_mapping), sep = ": ", collapse = "\n")
  legend_grob <- textGrob(legend_text, x = unit(0, "npc"), y = unit(1, "npc"), just = c("left", "top"), gp = gpar(fontsize = 7))

  # Combine the plots in a grid format with 3:2 width ratio for p2 and p1
  combined_plot <- (p2 + p1 + plot_layout(widths = c(3, 2))) / wrap_elements(legend_grob) + plot_layout(heights = c(2, 1))

  combined_plot
}

plot1 <- plot_FBID_clusters(obj, 'sctype_classification')

pdf(paste0(dir_save, "prop_FBID_celltype_res.clustree.pdf"), height = 5, width = 10)
print(plot1)
invisible(dev.off())
```











## Radial organization of scRNAseq clusters - to get around issue with differing cluster IDs - GIS - Ross something Florida

## Look at expression of our immune correlates in the LN after vaccination with M72 - spatial expression of different markers on different cell types to understand how M72 activates immune cells in the LN to generate memory - healthy participants vaccinated with M72 vs placebo - or potentially in the lung? Tissues sampled with biopsy / FNA

#### **Jaccard Index**

**(for overlap of DEGs)**

**What it does**: The Jaccard Index measures how much overlap there is between the DEGs identified in two different leave-one-out iterations (when you exclude one participant at a time). It tells you what percentage of genes are found in both iterations.

**How it helps**: A high Jaccard score means that most of the same DEGs are identified in both iterations, which suggests that your results are robust and consistent, even when you remove one participant. This shows that the DEGs arent being heavily influenced by any single participant.

```{r}

# Get the counts for each Mtb group within "No pathology"
  cell_counts <- obj@meta.data %>% #CHECK
    filter(Pathology == "No pathology") %>%
    select(PID, Mtb, Pathology) %>% 
    group_by(PID, Mtb, Pathology) %>%
    summarise(TotalCell_count = n(), .groups = 'drop')
  

# Function to calculate Jaccard Index
calculate_jaccard <- function(list1, list2) {
    intersection <- length(intersect(list1, list2))
    union <- length(union(list1, list2))
    if (union == 0) return(0)  # Avoid division by zero if both lists are empty
    return(intersection / union)
}

# Compute the Jaccard Index for each pair of leave-one-out results
num_participants <- length(participant_ids)
jaccard_matrix <- matrix(nrow = num_participants, ncol = num_participants,
                         dimnames = list(participant_ids, participant_ids))

for (i in 1:num_participants) {
    for (j in 1:num_participants) {
        genes_i <- gene_presence_count[[i]]
        genes_j <- gene_presence_count[[j]]
        jaccard_matrix[i, j] <- calculate_jaccard(genes_i, genes_j)
    }
}

# Visualize the Jaccard similarity matrix
heatmap(as.matrix(jaccard_matrix), Rowv = NA, Colv = NA, main = "Jaccard Similarity Matrix",
        scale = "none")

# Refined visualization of Jaccard similarity matrix
# Create a function to remove the lower triangle of the matrix
remove_lower_triangle <- function(mat) {
  mat[lower.tri(mat)] <- NA  # Set lower triangle to NA
  return(mat)
}

# Mask the lower triangle of the Jaccard similarity matrix
matrix_upper <- remove_lower_triangle(jaccard_matrix) #CHECK CORRECT MATRIC HERE

# Melt the matrix into long format for ggplot
matrix_melted <- reshape2::melt(matrix_upper, na.rm = TRUE)  # Remove NA values during melt

# Create the heatmap with ggplot2
p <- ggplot(matrix_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), color = "black") +  # Display values inside tiles
  scale_fill_gradient(low = "white", high = "firebrick", limits = c(0, 1)) +  # Color scale
  scale_x_discrete(position = "top") +
  theme_classic() +
  labs(title = "Jaccard Similarity Matrix", x = "Iteration 1", y = "Iteration 2") #+
  #theme(axis.text.x = element_text(angle = 45, hjust = 1))

cowplot::ggsave2(file.path(dir_save, 'Jaccard_matrix.png'), p,
                 units = 'cm',
                 width = 18,
                 height = 10)

# Permutation test for Jaccard Index significance
num_permutations <- 1000
all_genes <- unlist(gene_presence_count)  # Flatten all genes into a single vector
list_sizes <- sapply(gene_presence_count, length)  # Store original list sizes

permuted_jaccard_scores <- replicate(num_permutations, {
    shuffled_genes <- sample(all_genes)
    # Split them into permuted lists respecting original list sizes
    permuted_lists <- split(shuffled_genes, rep(seq_len(num_participants), times = list_sizes))
    
    # Calculate Jaccard Index for all pairs in this permutation
    all_scores <- combn(seq_along(permuted_lists), 2, function(idxs) {
        calculate_jaccard(permuted_lists[[idxs[1]]], permuted_lists[[idxs[2]]])
    }, simplify = TRUE)
    
    mean(all_scores)  # Return the mean Jaccard Index for this permutation
}, simplify = TRUE)  # Ensure the output is simplified to a numeric vector

# Calculate the observed average Jaccard Index (excluding diagonal self-comparisons)
observed_jaccard <- mean(jaccard_matrix[lower.tri(jaccard_matrix, diag = FALSE)])
median_jaccard <- median(jaccard_matrix[lower.tri(jaccard_matrix, diag = FALSE)])
p_value <- mean(permuted_jaccard_scores >= observed_jaccard)
print(paste("P-value for observed Jaccard Index being higher than chance:", p_value))


# Save jaccard results
# Set the directory where you want to save the output
output_file <- paste0(dir_save, "Jaccard_output.txt")

# Open a connection to the text file
sink(output_file)

# Print minimum and maximum of permuted scores and the observed Jaccard Index
print(paste("Min permuted Jaccard:", min(permuted_jaccard_scores)))
print(paste("Max permuted Jaccard:", max(permuted_jaccard_scores)))
print(paste("Observed Mean Jaccard:", observed_jaccard))
print(paste("Observed Median Jaccard:", median_jaccard))
print(paste("P-value for observed Jaccard Index being higher than chance:", p_value))

# Close the connection to the text file
sink()  # Stops writing to the file and goes back to printing in the console
```

#### **Concordance Index**

**(C-Index; for rank agreement)**

**What it does**: The C-Index checks how similar the rankings of the DEGs are between two iterations. It looks at how well the order of significance (like the p-values or fold changes) is preserved between two iterations.

**How it helps**: A high C-Index (closer to 1) means that the rankings of DEGs are similar between iterations, indicating that the relative importance of each gene is stable, even when you remove one participant. This again shows that your results are not dependent on the presence or absence of any one participant.

**Simple example**: Imagine that in one iteration, Gene X is ranked as the most significant, and in another iteration, its also ranked highly. A high C-Index means that the ranking of the genes stays the same (or very similar) across leave-one-out iterations.

**Why use both?**

Jaccard Index tells you if the same genes are found across different iterations.

C-Index tells you if the ranking order of these genes (how significant or important they are) stays consistent.

**In summary:**

If both the Jaccard Index and C-Index are high, your DEGs are reproducible across iterations, and your findings are not driven by any single participant. This demonstrates that your analysis is robust and the results can be trusted.

```{r}

# Function to calculate the Concordance Index between two ranked lists with potentially different sets of genes
calculate_concordance_index <- function(ranks1, ranks2, genes1, genes2) {
  # Find the intersection of genes between the two sets
  common_genes <- intersect(genes1, genes2)
  
  # Subset the ranks based on the common genes
  ranks1_common <- ranks1[match(common_genes, genes1)]
  ranks2_common <- ranks2[match(common_genes, genes2)]
  
  # Remove any NA values (in case of missing ranks for common genes)
  valid_indices <- complete.cases(ranks1_common, ranks2_common)
  ranks1_common <- ranks1_common[valid_indices]
  ranks2_common <- ranks2_common[valid_indices]
  
  n <- length(ranks1_common)
  concordant <- 0
  discordant <- 0
  
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      # Check if the order of pairs (i, j) is the same in both rankings
      if ((ranks1_common[i] < ranks1_common[j] && ranks2_common[i] < ranks2_common[j]) ||
          (ranks1_common[i] > ranks1_common[j] && ranks2_common[i] > ranks2_common[j])) {
        concordant <- concordant + 1
      } else if ((ranks1_common[i] < ranks1_common[j] && ranks2_common[i] > ranks2_common[j]) ||
                 (ranks1_common[i] > ranks1_common[j] && ranks2_common[i] < ranks2_common[j])) {
        discordant <- discordant + 1
      }
    }
  }
  
  # Handle case where all pairs are invalid or there are no valid pairs
  if (concordant + discordant == 0) {
    return(NA)  # Return NA if no valid comparisons are possible
  }
  
  # C-Index is the ratio of concordant pairs to all pairs
  c_index <- concordant / (concordant + discordant)
  return(c_index)
}

# List of gene names corresponding to each leave-one-out iteration
deg_gene_list <- lapply(leave_one_out_results, rownames)

# Rank the DEGs by p-value for each leave-one-out result
deg_ranks_list <- lapply(leave_one_out_results, function(res) {
  return(rank(-log10(res$p_val)))  # Rank by p-value, adjust to your preference
})

# Initialize a matrix to store the Concordance Index for each pair of iterations
num_iterations <- length(leave_one_out_results)
c_index_matrix <- matrix(NA, nrow = num_iterations, ncol = num_iterations, dimnames = list(participant_ids, participant_ids))

# Calculate the pairwise Concordance Index between each leave-one-out iteration
for (i in 1:num_iterations) {
  for (j in 1:num_iterations) {
    if (i != j) {
      # Get the ranks and genes for the two iterations
      ranks_i <- deg_ranks_list[[i]]
      ranks_j <- deg_ranks_list[[j]]
      genes_i <- deg_gene_list[[i]]
      genes_j <- deg_gene_list[[j]]
      
      # Calculate the Concordance Index between iteration i and j
      c_index_matrix[i, j] <- calculate_concordance_index(ranks_i, ranks_j, genes_i, genes_j)
    } else {
      # Self-comparison should give a perfect concordance (C-Index = 1)
      c_index_matrix[i, j] <- 1
    }
  }
}

# Visualize the Concordance Index matrix without scaling
heatmap(as.matrix(c_index_matrix), Rowv = NA, Colv = NA, main = "Concordance Index Between Iterations", scale = "none")


# Refined visualization of Concordance Index matrix
# Create a function to remove the lower triangle of the matrix
remove_lower_triangle <- function(mat) {
  mat[lower.tri(mat)] <- NA  # Set lower triangle to NA
  return(mat)
}

# Mask the lower triangle of the Concordance Index matrix
matrix_upper <- remove_lower_triangle(c_index_matrix) #CHECK CORRECT MATRIC HERE

# Melt the matrix into long format for ggplot
matrix_melted <- reshape2::melt(matrix_upper, na.rm = TRUE)  # Remove NA values during melt

# Create the heatmap with ggplot2
p <- ggplot(matrix_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), color = "black") +  # Display values inside tiles
  scale_fill_gradient(low = "white", high = "firebrick", limits = c(0, 1)) +  # Color scale
  scale_x_discrete(position = "top") +
  theme_classic() +
  labs(title = "Concordance Index Between Iterations", x = "Iteration 1", y = "Iteration 2") #+
  #theme(axis.text.x = element_text(angle = 45, hjust = 1))

cowplot::ggsave2(file.path(dir_save, 'ConcordanceIndex_matrix.png'), p,
                 units = 'cm',
                 width = 18,
                 height = 10)

```
